---
title: "6: Part 2 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Apply special cases of the GLM (ANOVA) to real datasets
2. Interpret and report the results of ANOVA in publication-style formats

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
install.packages("agricolae")
library(agricolae)   #you may need to install/update package "htmltools" before installing and loading agricolae

Litter <- read.csv("/home/guest/EDA-Fall2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is used when you have one categorical explanatory variable with more than two categories. In R, we can run the model with the function `lm` or `aov`, the latter of which will allow us to run post-hoc tests to determine pairwise differences.

For example: Does total litter biomass differ among sites at the Niwot Ridge LTER? This is really asking: is the mean of litter biomass different among sites?
```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

summary(Litter.Totals) #looking at the min, max, and mean of dry mass, there's a lot of variance. We want to see if this variance is due to different sites?

#check all value of plotID. All the different sites, and how many samples were taken from each site. You'll see it's unbalanced (not every site has the same value, but you can still use the 1-way ANOVA) 
summary(Litter.Totals$plotID)

# Test for normality. Running a shapiro test for all of the sites.
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
# shapiro.test() performs the Shapiro-Wilk test of the null that data follows a normal distribution

shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
#for the first site^^ you'll see that the p-value is less than 0.05, so you can reject the null hypothesis, meaning that the residuals at this site do NOT follow a normal distribution. 

shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])

#After running the shapiro test on all of the sites, you'll see that only two sites (46&57) have a p-value greater than 0.05, so only two follow a normal distribution


#results: reject null in all except two: NIWO_057 and NIWO_046
#but method is robust from small deviations from normal distribution
qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass)
#^^looking at qq normal plot will reinforce that the data do not follow a normal distribution 

# Test for equal variance
# ANOVA is robust against departures from equal variance.
# bartlett.test() performs Bartlett's test of the null that the variances in each of the groups (samples) are the same.
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID) #specify data (dry mass) and the level (plot id)

#results: reject null i.e. variances are not equal

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova) #specify dataframe, which is continuous/dependent variable, and which is categorical variable 
#looking at the ANOVA summary, Df(degrees of freedom) is the number of levels-1. The Sum sq is essential each value's deviation from the mean, summed. Mean Sq (Mean squared) is Sum sq - Df. The Pr(>F) is the p-value. Since the p-value is greater than 0.05, you reject the null (that the means are all the same). By rejecting the null, you are saying that there is variation in the means across levels/categories.
#results: reject null hypothesis i.e. difference between a pair of group means is statistically significant

# Format ANOVA as lm
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)
#lm also works to run a ANOVA test, the only difference is the output summary and table. The output of the aov function is a traditional ANOVA test output, while the lm gives an output for each categorical variable and shows the mean, standard error, t, and p-value for each category

# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
plot(Litter.Totals.anova2)


# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(Litter.Totals.anova) #the object of TukeyHSD is always the result of an anova test 
#^^the results is a long list of every possible pairing with a column for difference, lower, upper end points of confidence interval, and the p-value for the comparisons. A small diff suggests that the means are close to one another.

# Extract groupings for pairwise relationships that have a similar mean
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE) #object must be a result of an ANOVA test. Group = true is saying that you want the function to group levels with similar means. THe output is a series of rows of levels. Any levels with the same letter in the "group" column have similar means.

Litter.Totals.groups

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)

#^^label is allowing you to label each box plot inside the grid
```

What information does the output give us? How might we report this information in a report?

> 

## TWO-WAY ANOVA

### Main effects
A two-way ANOVA allows us to examine the effects of two categorical explanatory variables on a continuous response variable. 

For example: What if we wanted to know how litter masses varied across functional groups and land cover class?
```{r}
# Format as aov
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)
#how litter mass varies across functional group and nlcd class. In the summaries, you'll see one row for each categorical data


# Format as lm
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way)

```

### Interaction effects

We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an interaction effects two-way ANOVA, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.

```{r}
# Format as aov
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way3) #note that now, rather than functionalGroup + nlcdClass like we did in the original ANOVA, we're doing functionalGroup * nlcdClass to see the interaction among functional group and class on dry mass. This will result in a row showing functionalGroup:nlcdClass and it's p-value

# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.

```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3)
#if you run the Tukey HSD and the interaction is not significant, then interpret the difference based on your first two-way ANOVA without interaction effects.

Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass)) #Creating a variable that's a combination of functional group and land cover


Litter.test <- cbind(Litter,Litter.interaction)

Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction) #comparing the means of each of the 24 interaction levels we just created.

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups

# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = nlcdClass, color = functionalGroup)) +
  geom_boxplot()
print(Litter.anova.plot)

```

