---
title: "Assignment 3: Data Exploration"
author: "Megan McClaugherty"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

The completed exercise is due on Sept 30th.

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r setup, include=FALSE}

#install.packages("formatR")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=80),tidy=TRUE)
```


```{r}
getwd()
#library(tidyverse)
library(ggplot2)
Neonics<-read.csv("/home/guest/EDA-Fall2022/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter<-read.csv("/home/guest/EDA-Fall2022/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
```
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection 
Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. 
Neonicotinoids are a class of insecticides used widely in agriculture. The 
dataset that has been pulled includes all studies published on insects. Why 
might we be interested in the ecotoxicology of neonicotinoids on insects? 
Feel free to do a brief internet search if you feel you need more background 
information.

> Answer: Insects perform a wide variety of ecosystem services that directly 
benefit humans, particularly pollination and decomposition. They are also an 
integral part of food webs acting as both consumers and prey of other species. 
Unforeseen impacts of neonicotinoids on insects could therefore have severe 
consequences for resources that humans rely on.

3.  The Niwot Ridge litter and woody debris dataset was collected from the 
National Ecological Observatory Network, which collectively includes 81 aquatic 
and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample 
forest litter and woody debris, and we will focus on the Niwot Ridge long-term 
ecological research (LTER) station in Colorado. Why might we be interested in 
studying litter and woody debris that falls to the ground in forests? Feel free 
to do a brief internet search if you feel you need more background information.

> Answer: Leaf litter and woody debris in forests provide nutrients for soils 
as they decompose, help retain soil moisure, and also serve as important habitat 
for a variety of forest organisms. Changes in leaf litter and woody debris in an 
area could have far-reaching impacts and potentially provide information about 
the overall health of standing vegetation.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of 
salient information about the sampling methods here:

> Answer: 
 1.Samples are taken from sites with woody vegetation greater than 2m tall 
 2. The number and size of plots depends on the stature of the vegetation around
 the tower airshed. One litter trap pair (one ground trap and one elevated trap) 
 is deployed for every 400m2 plot area, however, the number and location of the 
 litter traps within each plot depends on the type of vegetation in that plot, 
 available space, spacing requirements, and tower airshed size.
 3. Ground traps are checked annually. Elevated traps in deciduous sites are 
 checked once every two weeks but this sampling ceases during dormant months. 
 Elevated traps in evergreen sites once every one or two months year-round.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)
```


6.  Using the `summary` function on the "Effect" column, determine the most 
common effects that are studied. Why might these effects specifically be of 
interest?

```{r}
summary(Neonics$Effect) 
#Using the summary function to find how how of each type of effect was observed

```

> Answer: These effects are of interest because it is useful to see how the 
insects responsed to exposure to the chemicals and to see how many experienced 
each type of effect.

7.  Using the `summary` function, determine the six most commonly studied 
species in the dataset (common name). What do these species have in common, and 
why might they be of interest over other insects? Feel free to do a brief 
internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name) #using the summary function to find the 
most commonly studied species in the dataset.

```

> Answer: The six most commonly used species were Honey Bees, Parasitic Wasps, 
Buff Tailed BumbleBee, Carniolan Honey Bees, Bumble Bees, and Italian Honey Bees. 
Many of these species are pollinators that provide valuable agricultural 
services for humans. If we use insecticides to target crop pests, we would want 
to understand the impacts on these pollinator species. Parasitic wasps target 
species that may be considered agricultural pests, so similarly we would want to 
understand how insecticides affect these species. 

8.  Concentrations are always a numeric value. What is the class of 
Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.) 
#Determining the class of the Conc.1..Author column.

```

> Answer: Conc.1..Author is a factor. They represent different categories of 
concentrations because these concentrations depend on several other factors, 
like the chemical type and concentration type, so they aren't universal 
concentrations.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by 
publication year.

```{r}
ggplot(Neonics)+
  geom_freqpoly(aes(x = Publication.Year), bins=50) 
  
#Using a frequency line graph to see the frequency of publications per year
```

10. Reproduce the same graph but now add a color aesthetic so that different 
Test.Location are displayed as different colors.

```{r}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year, color=Test.Location), bins=25) 

#Determining the number of publications per year by the location of each test/experiment
```

Interpret this graph. What are the most common test locations, and do they 
differ over time?

> Answer: The Lab and Field Natural are the most common test locations. This 
changes slightly over time. From the early 1990s into 2000, Field Natural was 
slightly more common but from that time forward, Lab test locations were 
relatively as common until about 2010 when Lab tests far exceeded all other test 
locations.

11. Create a bar graph of Endpoint counts. What are the two most common end 
points, and how are they defined? Consult the ECOTOX_CodeAppendix for more 
information.

```{r}

ggplot(Neonics)+
  geom_bar(aes(x=Endpoint))+
   scale_x_discrete(guide = guide_axis(n.dodge=2)) 
   
#using a bar graph to visualize the Endpoints coluumn counts
```

> Answer: The two most common endpoints are LOEL(Lowest Observable Effects Level), 
or the lowest concentration 
producing effects that were significantly different from the controls, and the 
NOEL (No Observable Effect Level), the highest concentration that produced 
effects that were not significantly different than those observed in controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date 
and confirm the new class of the variable. Using the `unique` function, 
determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate) 
#looking at how RStudio is interpreting the collectDate column.

library(lubridate)


Litter$collectDate<- as.Date(Litter$collectDate, format= "%Y-%m-%d") 
#using Lubridate to correct the collectDate column to a date format
class(Litter$collectDate)
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot 
Ridge. How is the information obtained from `unique` different from that 
obtained from `summary`?

```{r}
unique(Litter$plotID) 
#using the unique function to see how many plots were 
#sampled and comparing it to the output of the summary function
summary(Litter$plotID)
```

> Answer: 12 plots were sampled at Niwot Ridge. Unique only returns a list of 
values that don't have duplicates based on your selection criteria. Summary will 
give a count of how many entries there are for each of your selections.

14. Create a bar graph of functionalGroup counts. This shows you what type of 
litter is collected at the Niwot Ridge sites. Notice that litter types are fairly 
equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter)+
  geom_bar(aes(x=functionalGroup)) 
#creating a bar graph showing what types of litter are collected
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot 
of dryMass by functionalGroup.

```{r}
ggplot(Litter)+
  geom_boxplot(aes(x=functionalGroup, y=dryMass)) 
#using a bar graph to visualize the types and amounts of dryMass by functional 
#group in the dataset

ggplot(Litter)+
  geom_violin(aes(x=functionalGroup, y=dryMass))


```

Why is the boxplot a more effective visualization option than the violin plot in 
this case?

> Answer: In this case the boxplot is a more effective visualization option 
because without further manipulation, the violin plot is too compressed to reveal 
anything about the data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles, mixed, and twigs/branches have the highest biomass at these 
sites.
