---
title: "Assignment 09: Data Scraping"
author: Megan McClaugherty
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)
#1 Checking working directory, loading packages, setting theme.

getwd()

library(tidyverse)
library(rvest)
library(dplyr)
library(lubridate)


A9Theme <- theme_light(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(A9Theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 reading in the LWSP website for Durham in 2021

DurhamLWSPwebpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
DurhamLWSPwebpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3 Scraping the data for the 4 variables of interest

water.system.name <- DurhamLWSPwebpage %>% 
   html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name
  
pswid <- DurhamLWSPwebpage %>% 
   html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pswid

ownership <- DurhamLWSPwebpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership
  
max.withdrawals.mgd <- DurhamLWSPwebpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4 Creating a dataframe with the 4 variables
DurhamLWSPdf <- data.frame ("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr",    "Aug","Dec"),
                           "Water_System_Name" = (water.system.name),
                           "Ownership" = (ownership),
                           "PSWID" = (pswid),
                           "Max_Withdrawals_MGD" = as.numeric(max.withdrawals.mgd)) %>% 
                          mutate (Date = my(paste0(Month,"-",2021)))
                           
#arranging in chronological order
DurhamLWSPdf <- arrange(DurhamLWSPdf, Date)


#5 Line plot of max daily withdrawals for each month in 2021.
MaxWithdrawalbyMonth <- ggplot(DurhamLWSPdf, aes(x = Month, y = Max_Withdrawals_MGD, group = 1)) + 
  geom_line()+
   scale_x_discrete(limits = month.abb)+
  labs(title = "2021 Maximum Daily Withdrawals for Durham",
       y = "Maximum Withdrawal (MGD)")
print(MaxWithdrawalbyMonth)

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. Creating a scrape function to scrape data for any PWSID and year

baseURL <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'


scrape.it <- function(PWSID, Year){
        website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', PWSID, '&year=',      Year))
watersystem_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
pwsid_tag <- "td tr:nth-child(1) td:nth-child(5)"
ownership_tag <-"div+ table tr:nth-child(2) td:nth-child(4)"
max.withdrawals.mgd_tag <- "th~ td+ td"

#reading in the data for each of the 4 variables 
watersystem <- website %>% html_nodes(watersystem_tag) %>% html_text()
pwsid <- website %>% html_nodes(pwsid_tag) %>% html_text()
ownership <- website %>% html_nodes(ownership_tag) %>% html_text()
maxwithdrmgd <- website %>% html_nodes(max.withdrawals.mgd_tag) %>% html_text()

withdrawalsdf <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug","Dec"),
                             "Year" = rep(Year,12),
                             "Max.Daily.WithdrawalMGD" = as.numeric(maxwithdrmgd)) %>% 
  mutate(Watersystem = !!watersystem,
         PWSID = !!pwsid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#Setting the year to 2015 and creating the unique scrape url
Durham2015df <- scrape.it('03-32-010', 2015)

Durham2015df <- arrange(Durham2015df, Date)

#plotting 2015 daily max withdrawal data
Durham2015plot <- ggplot(Durham2015df, aes(x = Month, y = Max.Daily.WithdrawalMGD, group = 1))+
                  geom_line()+
                  scale_x_discrete(limits = month.abb)+
                  labs(title = 'Maximum Daily Withdrawals for Durham in 2015',
                       y = 'Maximum Daily Withdrawal (MGD)')
print(Durham2015plot)


```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 Using the function to extract data for Asheville in 2015.
Asheville2015df <- scrape.it('01-11-010', 2015)

Asheville2015df <- arrange(Asheville2015df, Date)


#Combining the two dataframes and creating a plot that compares the withdrawal of Asheville and Durham
AshevilleandDurham2015 <- bind_rows(Asheville2015df, Durham2015df)
AshevilleandDurhamplot <- ggplot(AshevilleandDurham2015, aes(x=Date, y=Max.Daily.WithdrawalMGD, 
                                                             color = Watersystem))+
                          geom_line()+
                          scale_x_date(date_breaks = "1 month",
                          date_labels = "%b")+
                          labs(title = "2015 Maximum Daily Withdrawals",
                          y = "Maximum Withdrawal (MGD)",
                          x = "Month")
                                                        
print(AshevilleandDurhamplot)
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9 Using the scape function I created to pull 2010-2019 data and plotting.

theyears <- rep(2010:2019)
Asheville_PWSID <- '01-11-010'

Asheville_NineYear <- map(theyears, scrape.it, PWSID=Asheville_PWSID)

Asheville_NineYear_df <- bind_rows(Asheville_NineYear)

Asheville_NineYear_plot <- ggplot(Asheville_NineYear_df, aes(x=Date, y=Max.Daily.WithdrawalMGD))+
                          geom_line() +
                          geom_smooth()+
                          labs(title = "Maximum Daily Withdrawals in Asheville",
                          y = "Maximum Withdrawal (MGD)",
                          x = "Year")

print(Asheville_NineYear_plot)

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
From 2009 to about 2017, it appears that Asheville's water usage remained relatively stable, but after around 2017 it appears to be trending upward.