---
title: "Assignment 5: Data Visualization"
author: Megan McClaugherty
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 Getting working directory, loading necessary packages and datasets

getwd()

library(tidyverse)
library(lubridate)
library(cowplot)
library(RColorBrewer)

LakeChemNutrient<-read.csv("/home/guest/EDA-Fall2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

LitterMassTrap<-read.csv("/home/guest/EDA-Fall2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 Formatting date

LakeChemNutrient$sampledate<-as.Date(LakeChemNutrient$sampledate, format = "%Y-%m-%d")
LitterMassTrap$collectDate<-as.Date(LitterMassTrap$collectDate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3 Building a theme and setting it as default theme to be applied across all plots

A5theme<-theme_light(base_size = 12)+
  theme(axis.text = element_text(color = "black"),
        axis.title.x = element_text(color = "black"),
  legend.position = "bottom")

theme_set(A5theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4 Using a scatterplot with a trend line to visualize the relationship between total phosphorus and phosphate concentrations, using different colors for the two lakes. Adjusted y-axis to remove an outlier.
PhosphatePhosphorus<-ggplot(LakeChemNutrient, aes(x=tp_ug, y=po4, color=lakename))+
  geom_point()+
  geom_smooth(method = lm, color="black")+
  ylim(0,45)+
  xlim(0,150)+
  xlab("Total Phosphorus Concentration (ug/L)")+
  ylab("Phosphate (ug/L)")

print(PhosphatePhosphorus)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5 Creating separate boxplots for temperature, total phosphorus, and total nitrogen by month for each lake.

LakeTempbyMonth<-ggplot(LakeChemNutrient, aes(x= factor(month, levels = c(1:12)), y=temperature_C, color = lakename))+
  geom_boxplot()+
  xlab("Month")+
  ylab("Temperature (C)")+
  scale_x_discrete(drop = FALSE)+
scale_color_brewer(palette = "Set1")
print(LakeTempbyMonth)

LakeTPbyMonth<-ggplot(LakeChemNutrient, aes(x= factor(month, c(1:12)), y = tp_ug, color = lakename))+
  geom_boxplot()+
  xlab("Month")+
  ylab("Total Phosphorus Concentration (ug/L)")+
  scale_color_brewer(palette = "Set1")+
  scale_x_discrete(drop = FALSE)

print(LakeTPbyMonth)

LakeTNbyMonth<-ggplot(LakeChemNutrient, aes(x = factor(month, c(1:12)), y=tn_ug, color = lakename))+
  geom_boxplot()+
  xlab("Month")+
  ylab("Total Nitrogen Concentration (ug/L)")+
  scale_color_brewer(palette = "Set1")+
  scale_x_discrete(drop = FALSE)

print(LakeTNbyMonth)

#cow plot to combine all three graphs into one output

LakeTempTPTNbyMonth <- plot_grid(LakeTempbyMonth + theme (legend.position = "none"), LakeTNbyMonth + theme (legend.position = "none"), LakeTPbyMonth + theme (legend.position = "bottom"), nrow = 3,  align = "v", rel_heights = c(2,2,2))

print(LakeTempTPTNbyMonth)

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: In the warmer months, especially July and August, the range and median temperature increases in both lakes compared to cooler months, with the highest median temperature occuring in September. For months with available data, the median temperature in Paul Lake is always slightly higher than Peter Lake except in October and November. For total phosphorus, the range and median increases relatively gradually from May through September in Peter Lake. However, in Paul Lake, the range and median decrease from May through July, then appears to stabilize before increeasing again in September. The median total phosphorus in Paul Lake is consistently lower than that in Peter Lake. While both have numerous high value, there is a broader distribution of outliers in Peter Lake than in Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6 
NeedleMass<-ggplot(subset(LitterMassTrap, functionalGroup == "Needles"),
                 aes(x = collectDate, y = dryMass, color = nlcdClass))+
                 xlab("Collect Date")+
                 ylab("Dry Mass (g)")+
                 geom_point()

print(NeedleMass)


#7
NeedleMassFacet<-ggplot(subset(LitterMassTrap, functionalGroup == "Needles"), 
             aes(x = collectDate, y = dryMass)) +
             xlab("Collect Date")+
             ylab("Dry Mass (g)")+
             geom_point() +
             facet_wrap(vars(nlcdClass), nrow = 3)
print(NeedleMassFacet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 seems more effective because it allows you to see a more comprehensive distribution of each NLCD class over time and easily make comparisons among the classes. In plot 6, there is a lot of overlap in the data, so even with color it is difficult to observe patterns or trends within or between the classes. 