---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Megan McClaugherty"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1 Checking working directory, loading packages, importing the dataset, and formatting the sample date column to date.
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
LakeChemistry<-read.csv("/home/guest/EDA-Fall2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
LakeChemistry$sampledate<-as.Date(LakeChemistry$sampledate, format = "%m/%d/%y")

#2 Setting a ggplot theme as the new default theme.
A6Theme<-mytheme <- theme_light(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature during July does not change with depth across all lakes.
Ha: Mean lake temperature during July changes with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Creating a new dataset for the variables of interest and omitting NAs for this new dataset.
JulyTemperature<-LakeChemistry %>% 
  mutate(month = month(sampledate)) %>% 
  filter(month == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

#5 Visualizing the relationship between temperature and depth using a scatterplot and adding a line of best fit. 

JulyTemperatureplot<-ggplot(JulyTemperature, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ylim(0, 35)+
  ylab("Temperature (C)") +
  xlab("Depth (m)")

print(JulyTemperatureplot)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as depth increases, temperature decreases. The distribution of points shows a negative linear relationship since they decrease rather consistently where there is the most data points.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 Using the lm function to test the linear relationship between temperature and depth.

TemperatureRegression <- lm(data = JulyTemperature, temperature_C ~ depth)
summary(TemperatureRegression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model has a p-value much lower than 0.05, which means that the null hypothesis should be rejected and that July temperature does change with depth in the lakes in this dataset. The R-squared is 0.7387, so 73.87%% of the variability in temperature is explained by changes in depth. The degrees of freedom on which this finding is based is 9726. Temperature is predicted to decrease by about 1.95 degrees C for every 1m change in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 Using an AIC to examine the explanatory variables and see which best predicts temperature. 

TempAIC<-lm(data = JulyTemperature, temperature_C ~ year4 + daynum +depth)
step(TempAIC)


#10 Running a multiple regression based on the variables year4, daynum, and depth as suggested by the AIC method.

TempMultipleRegression <- lm(data=JulyTemperature, temperature_C ~ year4 + daynum + depth)
summary(TempMultipleRegression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using year4, daynum, and depth to predict temperature because the AIC is lowest without removing any other these variables. This model explains 74.12% of the variation in temperature. This is an improvement over the simple linear regression, in which depth explained about 67% of the variation in temperature.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 Running the two types of ANOVA models to see if average lake temperature in July is different across lakes.

JulyTemperatureAnova<-aov(data = JulyTemperature, temperature_C ~ lakename)
summary(JulyTemperatureAnova)

JulyTemperatureAnovalm <- lm(data = JulyTemperature, temperature_C ~ lakename)
summary(JulyTemperatureAnovalm)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Since the p-value is very small and well below 0.05, this suggests there is a significant difference in mean temperature in the July among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
JulyTemperatureplot2 <- ggplot(JulyTemperature, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5) +
  guides(color = guide_legend(title = "Lake Name"))+
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 35) +
  ylab("Temperature (C)") +
  xlab("Depth (m)")

print(JulyTemperatureplot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Using the Tukey's HSD test on the July Temperature ANOVA model to determine the pairs of lakes with mean July temperatures that are most related.

TukeyHSD(JulyTemperatureAnova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake statistically has the same temperature as Peter Lake, since the difference is very small. Ward Lake is also very similar. Central Long Lake is statistically distinct from all of the other lakes because all of the pairings for that lake have a high difference in means. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were only interested in looking at Peter and Paul Lakes, we could use a two sample t test to determine if their means are distinct.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#Filtering for only Crampton Lake and Ward Lake
JulyTempCramptonWard <- JulyTemperature %>% 
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")
#Running the two sample t-test to see if there are differences in the mean July temperature for the two lakes.
JulyTempCramptonWardTtest <- t.test(JulyTempCramptonWard$temperature_C ~ JulyTempCramptonWard$lakename)
JulyTempCramptonWardTtest


```

>Answer: The p-value is about 0.3 suggesting that the null hypothesis is true, that there is not a statistically signifcant difference between the means in these two lakes. While they are not equal they are statistically very similar, which reflects the very small difference of about -0.89 found in part 16.
